# Getting Started {#start}


```{block, type = "los"}

You can skip this chapter if you can:

  * Use the console to input simple lines of code

  * Install and load a package

  * Recognise a function 

```

To use this chapter you will need a working version of R and R Studio.

If you can open R Studio and type `2 + 2` in the **console** you are good to go. 

If that is confusing, you might need to jump back to [installing R Studio](#install_rs) and the section [navigating R Studio](#navigate_rs).


## The Big Secret {#thesecret}
Okay. This is the Big Secret. The thing you will not believe. The most important thing you'll learn about R. 


```{block, type = "hey"}

You should copy and paste other peoples' code. 

```


Yes, I'm serious. You should copy and paste and edit code you find online^[If you're reading this as part of your coursework you might be panicking about plagiarism here, after all, we spend a lot of time telling you plagiarism is the worst thing you could ever do and that we'll use software to detect it. Code is a bit different. We are always trying to find the most efficient way of doing something, and so ideally you would all write code that was identical. Sadly, humans naturally differ in the way the think about problems. My job would be a lot easier if everybody thought the same. If I have set you this book as reading, I can swear to you I will never put your code through a plagiarism checker. That would be very stupid.]. In this book you should always be copying bits of code and pasting it into your console. That is why I publish the code alongside it. 

It took me about ten years of working in R to get over my fear of copying and pasting other peoples' code. Now I even have a [repository](https://github.com/jillymackay/commoncode) of all the code I use over and over so I can copy and paste my own code.

Most people will not remember code off the top of their head. This is why we use the help function (try typing `?summary` to see what it says) to look at the documentation, and resources like Stack Exchange and Google to help us (more in [troubleshooting](#trouble)).


### Exercise {#ex_tidytuesday}
You might be very skeptical here about how much I want you to be copying code. If you're not convinced, I suggest you go watch my RStats Crush the amazing [David Robinson](https://twitter.com/drob) do one of his Tidy Tuesday screencasts. David Robinson has forgotten more R than I'll ever learn, and see - he still [copies and pastes code](https://www.youtube.com/channel/UCeiiqmVK07qhY-wvg3IZiZQ).


## Projects {#projects}
I would like to begin by creating a new project in R Studio.

* Open R Studio

* Go to `File > New Project`

* Set up a new project either in an existing directory (folder), or in a new one. Ignore repositories for now. 


You can call this project anything you like. I recommend something short like 'start'.

A R Project sits inside a folder. Everything inside that folder is part of that R Project. Let's say you have a folder called:

`mydata`

and you create an R Project called `analysis`.

The `analysis.Rproj` will live inside the `mydata` folder, and it will 'see' all the data, files, and any other folders, you have in `mydata`. 

Using R Studio you will get a lot better at 'directory management', or knowing where you have saved things. 


## Environment {#environment}
In your brand new project you will have a clean environment (refer back to [Figure 2.5](#navigate_rs)). Our first exercise is going to explore what the environment really means. 

### Exercise - Environment {#ex_env}
When we first installed R, you were typing mathematical equations into the console. Let's remind ourselves of this by running an equation now. In the console, type a long and complicated equation, then hit 'enter' when you're done.

```{r}

500 + 23 / 91

```


If this was a calculation we often ran, we might want to save the answer. We can do this by using the **assign** symbol **<-**. 

For example, if we type this:

```{r}

x <- 500 + 23 / 91

```

... you will see that we simply get a new line returned. We don't get the answer. However, in your environment you will see under the heading 'Values', a new thing called `x`. 

Enter the following code:

```{r}
x

```

R remembers what x is and recalled it. This is useful if we want to do more with `x` after we've calculated it. 

```{r}

x /100


```

As well as values, we can ask R to remember a string of text as well. Try this:


```{r}

hello <- "Computers always print 'Hello World!' but we never say hello back"


print(hello)

```

But R is very fussy. What happens if you run this code:


```{r eval = FALSE}

print(HELLO)

```

You see, R ise **case sensitive**, meaning we have to be careful to always type things the same. This is one of the reasons why R Studio is really handy, because you can see what you've saved in your environment. 


Now type this:

```{r}

hello <- "Hello, R!"

```

Take a guess what is going to happen before you type the next line. Were you right?

```{r}

print(hello)

```



### Exercise - Functions {#ex_func}
#### The Basic Function

A function is a handy way to bundle together some lines of code. David Robinson [says that](https://twitter.com/drob/status/928447584712253440?s=20) if you run a few lines more than twice you should write a function to do it instead. 

I'm going to contradict myself now. Functions are really important in R, but I'm not going to ask you to write one anywhere else but in this exercise. Most things in this book I expect you to practice over and over, but functions are useful to understand [packages](#packages), and that's why we're going to talk about them now.  

Let's create a function to make R welcome us. Copy and paste this code into a new script file (you may want to save it as your functions example), and then run it. 

```{r}
mynameis <- function (name) {
  print (paste0("Hello, ", name, ", how are you today?"))
}
mynameis ("Jill")
```

**What is this code saying?**

```{block, type = "translate"}

Create a new thing called `mynameis`

Give (`<-`) `mynameis` the `function` purpose. Everything inside the curly brackets `{}` is the function. 

The `function` should make a new thing called `name`.

The `function` should `print` the result of pasting a string (`paste0`) of things, the words `Hello, `, whatever you said `name` should be, and `, how are you today?`. 


```



You may have noticed that, just like when we were assigning `x` a value, we now have something in our environment called `mynameis`.

Whenever we type `mynameis("Jill")` into the console, or execute that line in a script file, R knows it should look to the environment and run the code that's been bundled up into that package. 

For example, we could make R welcome us in exactly the same way by typing:

```{r}
print (paste0("Hello, ", "Jill", ", how are you today?"))

```

But we would need to change the name every time we wanted welcome a new user. **We are lazy, so we use the function to reduce the amount of things we need to change to get a different outcome.** 

Functions make our code more standardised. If we all define the `mynameis` function at the start of our documentation, we don't need to worry about accidentally deleting an important character in the code. This is especially important as we like to copy-paste code when we need to. 

Most scientists will probably not write their own functions, but you should know how they work.



## Packages {#packages}
If people write very good and useful functions, they usually want to share them. They can do this with packages. 

There are packages for everything, packages for drawing maps, packages for analysing data, packages for suggesting how to analyse your data, packages to give you more data, packages for tidying your data . . . this is one of the reasons we might all have different bits of code. There can be lots of ways to do something. 

### Exercise - Install and Load a Package {#ex_packages}

Packages work in two stages. The very first time you need a package you need to install it. Then, every time you start R you need to load the package into your livrary.

To install a package you need to type `install.package("package_name")`. 

To load it into your library you need to type `library(package_name)`.

When you load a package from your library it goes into a special hidden environment where you can access all the functions that the package has written for you.

For example, we will make great use of the `tidyverse` package in this book. You need to start by installing it with:

```{r eval = FALSE}

install.packages("tidyverse")

```

And when it has finished downloading, you need to type:

```{r eval = FALSE}

library(tidyverse)

```

to load it. You'll see that nothing has appeared in your Environment, but actually you have a whole host of new functions & data you can play with. For example:


```{r eval = FALSE}

slice(starwars)
```

Which should give you this:

```{r echo = FALSE, warning=FALSE, message = FALSE}

dplyr::slice(starwars)

```

Now we have a new dataset (`starwars`) and a new thing we can do (`slice` the top of the data). We will talk much more about `tidyverse` in the [data chapter](#data).


```{block, type = "hey"}

You will often get warnings in R when you type these code blocks. To keep the book tidy, I have suppressed (hidden) the warnings in this book. Don't worry about them - they're usually just trying to keep packages compatibile with older versions of R  


```

## R Markdown vs R Scripts {#scripts_rmd}

In R Studio you can use R Script files and R Markdown files to save your work. Both of these files can be created, edited, saved, and opened in your R Project file. You can send R Script and R Markdown files to other people, so they can edit and share your code too. 

If you are reading this book as part of coursework I've set you, you will likely have an assignment where you neeed to download and edit an R Markdown file, then send it back to me. 

### R Scripts
R script files are great for testing out small chunks of code on their own. 

Go to `File > New File > R Script` or press `Ctrl + Shift + N` to open one. 

Everything you type in an R script file is executable code. 

Unlike working in the **console**, you can type multiple lines of code in an R script, and run each one by hitting `ctrl + enter`.

You **can't type** plain text in an R Script file. Text has to be annotated by the comment symbol (`#`)


```{r, eval = FALSE, warning = FALSE, message = FALSE}

# In an R script file - you can write comments
# Which always begin with a hashtag

# Comments can be useful to say things like what date you wrote the script on,
# or what you were trying to do. 

print("You can also use script files to write multiple lines of code")

print("and edit those lines of code")


# before you run them by pressing ctrl + enter at the same time  when your cursor is on that line. 


```




### R Markdown
R Markdown allows you to combine code with actual text. This book was written in R Markdown


In R Markdown you can have a text chunk like this one.

And follow that with code, bracketed with three ticks like this:
\```{r}

print("followed by code like this")

\```

```{r echo = FALSE, warning = FALSE, message = FALSE}
print("followed by code like this")
```


[There is a great R Markdown cheatsheet from R Studio here](https://rstudio.com/wp-content/uploads/2015/02/rmarkdown-cheatsheet.pdf). I had this printed out and hung above my desk for years. 

### A video on R Markdown {#video_rmd}
If you would like to watch me using R Markdown, you're in luck! Theres [a video here](https://media.ed.ac.uk/media/R+ConversationsA+Intro+to+R+Markdown/1_td0q33v8/112983051).




## Video Introductions {#video_env}

There is a short introduction to R and R Studio in Video format here:

* [Video 1](https://media.ed.ac.uk/media/R+ConversationsA+Intro+to+R+Studio+1/1_aox3in51/112983051)

* [Video 2](https://media.ed.ac.uk/media/R+ConversationsA+Intro+to+R+Studio+2/1_vm1bylon)
